\documentclass{article}
\usepackage{amsmath, amssymb, amsthm}
\usepackage[margin=1in]{geometry}

\newcommand{\E}{\mathbb{E}}
\newcommand{\Prob}{\mathbb{P}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\ind}{\mathbf{1}}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{conjecture}{Conjecture}

\title{MDP for Survival Time Optimization}
\author{}
\date{}

\begin{document}
\maketitle

\section{Setup}

We optimize \textbf{expected survival time}, not lap count.

\begin{definition}[State]
After completing a lap, the particle is at gate phase $\Omega \in [0, P)$ where $P = T_1 + T_2$ is the gate period. A fresh lifespan $L \sim \text{Exp}(\lambda)$ is drawn.
\end{definition}

\begin{definition}[Gate]
The gate is ON during $\Omega \in (0, T_1]$ and OFF during $\Omega \in (T_1, P]$.
Laps only begin when the gate is ON, so $\Omega \in (0, T_1]$ initially.
\end{definition}

\begin{definition}[Survival Time Accumulation]
If the particle waits $w$ then travels $\tau$:
\begin{itemize}
    \item Time alive this lap attempt: $\min(L, w + \tau)$
    \item Survives iff $L \geq w + \tau$ AND gate is ON at arrival
\end{itemize}
\end{definition}

\section{Value Function for Survival Time}

\begin{definition}[Value Function]
$W(\Omega)$ = expected total survival time starting from phase $\Omega$ with a fresh lifespan draw.
\end{definition}

For a particle at phase $\Omega$ that draws lifespan $L$:
\begin{itemize}
    \item It chooses some policy (wait time, or sequence of laps)
    \item It accumulates survival time until death
    \item $W(\Omega) = \E_L[\text{survival} \mid \Omega, \text{optimal policy}]$
\end{itemize}

\section{Two Candidate Strategies}

\subsection{Strategy A: Wait (Late)}

Wait as long as possible, then do one lap arriving at the end of an ON window.

Given $(\Omega, L)$:
\begin{itemize}
    \item Target arrival at $\Omega' = T_1$ (end of ON window)
    \item Required wait: $w = T_1 - \Omega - \tau + kP$ for some $k \geq 0$
    \item Choose largest $k$ such that $w + \tau \leq L$
\end{itemize}

If successful:
\begin{itemize}
    \item Survival contribution: $w + \tau$ (possibly close to $L$)
    \item Continuation: $W(T_1)$
\end{itemize}

If no ON window reachable:
\begin{itemize}
    \item Survival contribution: $L$
    \item Dead
\end{itemize}

\subsection{Strategy B: Spam (Early)}

Do laps as fast as possible to ``re-roll'' the lifespan.

Given $(\Omega, L)$:
\begin{itemize}
    \item Find minimum wait $w_{\min}$ to reach next ON window
    \item Do lap, arrive at $\Omega'$, get new $L' \sim \text{Exp}(\lambda)$
    \item Repeat until death
\end{itemize}

Value of spamming:
\begin{itemize}
    \item Each successful lap contributes $w_{\min} + \tau$ to survival
    \item After $k$ laps, at phase $\Omega_k$ with fresh lifespan
    \item Expected survival = $\sum_{i=1}^{k}(w_i + \tau) + W(\Omega_k)$ on success path
\end{itemize}

\section{The Key Tradeoff}

\textbf{Waiting} maximizes survival \emph{conditional on this lifespan}.

\textbf{Spamming} tries to reach a state with potentially better future value.

The question: when is the expected value of spamming higher?

\section{Bellman Equation}

Let $V(\Omega, L)$ = expected survival from state $(\Omega, L)$ under optimal policy.

\[
V(\Omega, L) = \max_{\text{policies}} \E[\text{survival} \mid \Omega, L, \text{policy}]
\]

By memorylessness, define:
\[
W(\Omega) = \E_{L \sim \text{Exp}(\lambda)}[V(\Omega, L)]
\]

For a single-lap decision with wait time $w$:
\begin{itemize}
    \item Arrival phase: $\Omega' = (\Omega + w + \tau) \mod P$
    \item Gate open at arrival: $\Omega' \in (0, T_1]$
    \item Survival contribution: $\min(L, w + \tau)$
\end{itemize}

\[
V(\Omega, L) = \max_{w \geq 0} \left\{
\min(L, w + \tau) + \ind[L \geq w + \tau] \cdot \ind[\Omega' \in (0, T_1]] \cdot W(\Omega')
\right\}
\]

\section{Simplification: Two Actions}

Rather than optimizing over continuous $w$, consider two discrete actions:
\begin{enumerate}
    \item \textbf{Early}: Wait minimum time to reach start of next ON window
    \item \textbf{Late}: Wait maximum time to reach end of an ON window before death
\end{enumerate}

Let $w_E(\Omega)$ = minimum wait from $\Omega$ to reach an ON window.
Let $w_L(\Omega, L)$ = maximum wait from $\Omega$ such that $w + \tau \leq L$ and arrival is in ON window.

\subsection{Value of Early}

\[
V_E(\Omega, L) = \begin{cases}
(w_E + \tau) + W(\Omega_E) & \text{if } L \geq w_E + \tau \\
L & \text{otherwise}
\end{cases}
\]

where $\Omega_E = (\Omega + w_E + \tau) \mod P$ is the arrival phase (near start of ON window).

\subsection{Value of Late}

\[
V_L(\Omega, L) = \begin{cases}
L + W(T_1) & \text{if } L \geq w_L + \tau \text{ for some valid } w_L \\
L & \text{otherwise (wait until death)}
\end{cases}
\]

Wait—this isn't quite right. If late is possible, the survival contribution is $w_L + \tau$, which is close to but not equal to $L$ (it's the largest feasible wait+travel).

Let me reconsider...

\subsection{Corrected Value of Late}

If we can reach an ON window with late strategy:
\begin{itemize}
    \item Choose $w_L$ maximal such that $w_L + \tau \leq L$ and $(\Omega + w_L + \tau) \mod P \in (0, T_1]$
    \item Survival contribution: $w_L + \tau$
    \item Continuation: $W(T_1)$ (since we arrive at end of ON window)
\end{itemize}

\[
V_L(\Omega, L) = \begin{cases}
(w_L + \tau) + W(T_1) & \text{if reachable} \\
L & \text{otherwise}
\end{cases}
\]

\section{When is Early Better?}

Early is better than Late when:
\[
V_E(\Omega, L) > V_L(\Omega, L)
\]

Case 1: Both can complete a lap.
\[
(w_E + \tau) + W(\Omega_E) > (w_L + \tau) + W(T_1)
\]
\[
W(\Omega_E) - W(T_1) > w_L - w_E
\]

The LHS is the advantage of arriving at $\Omega_E$ vs $T_1$.
The RHS is the extra survival time from waiting longer.

If $W$ is monotonic, this gives a threshold condition.

Case 2: Only Early can complete a lap (Late would die).

This happens when $L < w_L^{\min} + \tau$ but $L \geq w_E + \tau$.

Then Early gets $(w_E + \tau) + W(\Omega_E)$ and Late gets $L$.

Early is better iff:
\[
(w_E + \tau) + W(\Omega_E) > L
\]

\section{The Spam Value}

What if instead of one Early lap, we do $k$ Early laps?

Expected value of $k$-lap spam starting from $(\Omega, L)$:
\[
V_{\text{spam}}^{(k)}(\Omega, L) = \E\left[\sum_{i=1}^{k} (w_i + \tau) \cdot \ind[\text{survived } i \text{ laps}] + W(\Omega_k) \cdot \ind[\text{survived } k \text{ laps}]\right]
\]

This is complex because each lap depends on the new lifespan draw.

\subsection{Expected Laps from Spamming}

Let $N(\Omega)$ = expected number of laps completed by spamming from phase $\Omega$.

$N(\Omega)$ satisfies:
\[
N(\Omega) = \Prob(L \geq w_E + \tau) \cdot (1 + N(\Omega_E))
\]

Let $p = \Prob(L \geq w_E + \tau) = e^{-\lambda(w_E + \tau)}$.

If spamming always lands at the same phase $\Omega_E$ (which happens if $w_E + \tau$ is a multiple of $P$... not generally true), then:
\[
N(\Omega) = p(1 + N(\Omega_E))
\]

In general, the phase evolves, making this a Markov chain on $[0, P)$.

\section{Conjecture: Threshold Structure}

\begin{conjecture}
There exists a threshold $c^*(\Omega)$ such that:
\begin{itemize}
    \item If $L < c^*(\Omega)$: Spam (go early, try to re-roll)
    \item If $L \geq c^*(\Omega)$: Wait (go late, maximize current life)
\end{itemize}

Moreover, $c^*(\Omega)$ may be constant (independent of $\Omega$) under certain symmetry conditions.
\end{conjecture}

\textbf{Intuition}: 
\begin{itemize}
    \item Small $L$: You're going to die soon anyway. Might as well try for a re-roll.
    \item Large $L$: You have runway. Don't waste it—wait as long as possible.
\end{itemize}

The threshold $c^*$ balances:
\begin{itemize}
    \item Value of re-roll: $W(\Omega_E)$
    \item Cost of re-roll: Lost waiting time ($w_L - w_E$)
\end{itemize}

\section{Solving for $W(\Omega)$}

Under the threshold policy with threshold $c$:
\[
W(\Omega) = \E_L[V(\Omega, L; c)]
\]

where $V(\Omega, L; c) = V_E(\Omega, L)$ if $L < c$ and $V(\Omega, L; c) = V_L(\Omega, L)$ if $L \geq c$.

This gives a fixed-point equation for $W(\Omega)$ as a function of $c$.

The optimal $c^*$ maximizes $W(\Omega)$.

\section{Warm-up: No Gate}

As a sanity check, consider the case with no gate (gate always open).

Every lap takes exactly $\tau$. No waiting. The value function $W$ satisfies:
\[
W = \E_L[\min(L, \tau) + \ind[L \geq \tau] \cdot W]
\]

Compute $\E[\min(L, \tau)]$ for $L \sim \text{Exp}(\lambda)$:
\begin{align*}
\E[\min(L, \tau)] &= \int_0^\tau L \lambda e^{-\lambda L} dL + \tau \cdot \Prob(L \geq \tau) \\
&= \frac{1}{\lambda}(1 - e^{-\lambda\tau}) - \tau e^{-\lambda\tau} + \tau e^{-\lambda\tau} \\
&= \frac{1}{\lambda}(1 - e^{-\lambda\tau})
\end{align*}

So:
\[
W = \frac{1}{\lambda}(1 - e^{-\lambda\tau}) + e^{-\lambda\tau} W
\]
\[
W(1 - e^{-\lambda\tau}) = \frac{1}{\lambda}(1 - e^{-\lambda\tau})
\]
\[
W = \frac{1}{\lambda} = \mu
\]

\textbf{Result:} Without a gate, expected survival time equals the mean lifespan $\mu$. This matches intuition—on average, you live exactly as long as your expected lifespan.

\section{With Gate: Fixed Point Equation}

With a gate, let's derive $W(\Omega)$ under a threshold policy.

Assume threshold $c$: spam if $L < c$, wait if $L \geq c$.

\textbf{Case: $L \geq c$ (Wait/Late)}

Target the end of an ON window. Let $w_L(\Omega, L)$ be the wait time to arrive at $T_1$.

If we can reach an ON window (i.e., $\exists k \geq 0$ with $w_L + \tau \leq L$):
\begin{itemize}
    \item Survival: $w_L + \tau$
    \item Continuation: $W(T_1)$
\end{itemize}

Otherwise (lifespan too short to reach any window):
\begin{itemize}
    \item Survival: $L$
    \item Dead
\end{itemize}

\textbf{Case: $L < c$ (Spam/Early)}

Do lap as fast as possible. Let $w_E(\Omega)$ be minimum wait to reach an ON window.

If $L \geq w_E + \tau$:
\begin{itemize}
    \item Survival: $w_E + \tau$
    \item Continuation: $W(\Omega_E)$ where $\Omega_E = (\Omega + w_E + \tau) \mod P$
\end{itemize}

Otherwise:
\begin{itemize}
    \item Survival: $L$
    \item Dead
\end{itemize}

\subsection{Fixed Point}

\[
W(\Omega) = \underbrace{\int_0^c V_E(\Omega, L) \lambda e^{-\lambda L} dL}_{\text{spam regime}} + \underbrace{\int_c^\infty V_L(\Omega, L) \lambda e^{-\lambda L} dL}_{\text{wait regime}}
\]

This is an integral equation in $W$. The solution depends on $c$.

\section{Finding Optimal $c^*$}

The optimal threshold $c^*(\Omega)$ satisfies:
\[
\frac{d}{dc} W(\Omega; c) = 0
\]

At $L = c$, the particle is indifferent between spam and wait:
\[
V_E(\Omega, c) = V_L(\Omega, c)
\]

This gives:
\[
(w_E + \tau) + W(\Omega_E) = (w_L + \tau) + W(T_1)
\]
\[
W(\Omega_E) - W(T_1) = w_L - w_E
\]

\textbf{Interpretation:} The gain in future value from arriving at $\Omega_E$ instead of $T_1$ must equal the lost waiting time.

\section{Special Case: $\Omega_E = T_1$}

If early and late both arrive at $T_1$ (e.g., symmetric gate timing), then:
\[
W(T_1) - W(T_1) = w_L - w_E
\]
\[
0 = w_L - w_E
\]

This is only satisfied if $w_L = w_E$, meaning early and late are the same strategy.

In general, $\Omega_E \neq T_1$, so the threshold is non-trivial.

\section{Hypothesis: One Discontinuity}

The user's intuition: there should be ``one discontinuity''—a single threshold $c^*$.

\textbf{Why this might be true:}
\begin{itemize}
    \item $V_E(\Omega, L)$ is linear in $L$ (for small $L$, you die; for larger $L$, you complete a lap)
    \item $V_L(\Omega, L)$ is also piecewise linear in $L$
    \item The crossing point defines $c^*$
\end{itemize}

\textbf{Potential complication:}
\begin{itemize}
    \item With multiple ON windows reachable, there might be multiple crossing points
    \item But likely only one is optimal (the first crossing from below?)
\end{itemize}

\section{Next Steps}

\begin{enumerate}
    \item Compute $W(\Omega)$ for a threshold policy (closed form?)
    \item Optimize over $c$ to find $c^*$
    \item Check if $c^*$ is independent of $\Omega$
    \item Verify against simulation
    \item If $c^*$ has a clean form, prove optimality of threshold policy class
\end{enumerate}

\end{document}
